On the Programmatic Generation of Reproducible Documents

### 2020-10-17 inst/jss/pre-review-response.md

# Pre-comment response

We would like to thank the editors for their comments and suggestions. We have made considerable changes to both the submission as well as the package based on the feedback provided. Below we address and discuss each of the issues provided.

# Response to the general comments from the editorial team

We have restructured the document and added text describing more specifically how a computational component list's structure maps to the that of an R Markdown document. We have also added an example of the `ld_cc_dendro()` function earlier in the document to illustrate the correspondence between the list and the document.

The progression of the examples has also been revised so that the example builds progressively in complexity and sophistication. We hope that this change allows the reader to digest successively more elaborate concepts and functionality facilitated by the package. However, we would like to remind the reviewers that the goal of the package is to address the creation of the presentation of results in the context described. It is not meant to address how those results are derived. We have added emphasis to this in Section 5.

The advantages and limitations of both the automated reproducible document generation and package are given in Section 1 and the discussion in Section 6. Section 5 and 6 describe those situations for which the package has been shown useful.

The comment concerning the "monolithicness" of the presentation is somewhat difficult to understand. The package itself adrresses the problem creating reproducible documents via lists and, in this sense, the task is singular and perhaps this is the motivation for the comment. However, the conceptual contribution is the separation of narrtive from computational components along with the analysis and dicussion of this characterization and it's consequences. The technical contribution is the implementation of this capability in a way that is general, flexible, and easy to integrate into data processing pipelines. An earlier version of the manuscript was sufficient for analysts (intermediate R users) to create reproducible documents for reporting to clinicians with little to no extra instruction. The manuscript should be sufficent for the reader to understand the underlying concepts, understand how to automate R Markdown document generation for a variety of different types of documents and document types, and extend the use of the package to the users own needs. The paper is not meant to show readers how to create complete reports for a specific domain and the clinical trial example is only meant to illustrate one outlet for the package, which we have found useful. If there are specific suggestions for how else to address this issue, we would be happy to incorporate them.

# Further comments

The manuscript needs to be spell-checked and proofread. The manuscript in its current form is not suitable for review.

- Hopefully it is in the current form.

A print method should be added for listdown objects.

- This is included in the newest revision of the package and is described in the paper.

As replication material an R script, e.g., by purling the Rmd file and  editing the R script output such that it is in a human readable form, needs to be provided.

- This is included in the updated `supplemental-materials` directory.

The submitted package version and the version on CRAN are out of sync. An updated version should be submitted to JSS.

- We have uploaded version 0.2.21 to CRAN and included it in this submission.

Appendix 1 can be omitted, but rather included in plain text form in a standalone file in the supplementary material.

- Done, thanks for the suggestion.

For referring to subsections, do not use Subsection x.y, just Section x.y.

- This has been done. Please note that we do describe creating "subsections" and "subsubsection" in the text and that should be distinguished from the references.

The code presented in the manuscript should not contain comments within the verbatim code. Instead the comments should be made in the normal LaTeX text.

- Done.

For the code layout in R publications, we typically distinguish input/output using Sinput/Soutput (or equivalently CodeInput/CodeOutput). Unless there are special reasons to format it differently, the input should use the text width (up to 76 or 77 characters) and be indented by two spaces, e.g.,

- Done.

As a reminder, please make sure that \proglang, \pkg and \code have been used for highlighting throughout the paper (including titles and references), except where explicitly escaped.

- Done.

Springer-Verlag (not: Springer) ysis.” In Compstat, pp. 575–580. Springer.

- Done. Thanks for pointing this out.

Please make sure that all software packages are cite'd properly.

- Done.

All references should be in title style.

- Done

Please make sure that the files needed to replicate all code/examples within the manuscript are included in a standalone replication script

- Done

### 2020-10-17 inst/jss/supplemental-material/listdown-jss.tex
Decorators can take the returns of functions.

\author{
Michael Kane\\Yale University \And Xun (Tony) Jiang\\Amgen
Inc \And Simon Urbanek\\The University of Aukland
}
\title{On the Programmatic Generation of Reproducible Documents}

@@

      Xun (Tony) Jiang\\
  Amgen Inc\\
  One Amgen Center DriveThousand Oaks, CA 91320-1799, USA\\
  E-mail: \href{mailto:xunj@amgen.com}{\nolinkurl{xunj@amgen.com}}\\

@@

One area where we have found \pkg{listdown} is particularly useful is in
the reporting and research of clinical trial data. These collaborations
tend to be between (bio)statisticians and clinicians either analyzing
past trial data to formulate a new trial or in trial monitoring where
trial telemetry (enrollment, responses, etc.) is reported and initial
analyses are conveyed to a clinician. The associated presentations
require very little context - clinicians often have as good an
understanding of the data collected as that of the statistician's -
often eliminating or significantly reducing the need for narrative
components. At the same time, a large number of hierarchical,
heterogenous artifacts (tables and multiple types of plots) need to be
conveyed thereby making the manual creation of \proglang{R} Markdown
documents inconvenient.

In this case, data presentation can be fixed across trials. This is
especially true in the initial stages, which focus on patient
demographics and enrollment. This approach has made it convenient for
our group to quickly generate standardize and complete reports for
multiple trials concurrently. To date, we have used listdown to report
on five clinical trials, with another two currently in process. Results
are disseminated using the \pkg{workflowr} package, usually with nine
tabs conveying aspects of the data from collection through several
different analyses, and each tab containing approxiately five to thirty
tables, plots, or other artifacts including \pkg{trelliscopejs}
\citep{trelliscopejs} environments which may hold hundreds of graphs. By
generating many presentation artifacts we are able to address data-drive
questions and issues during collaborative sessions and by carefully
structuring these elements we allowing all members to participate in the
process.

@@

\begin{itemize}
\item{\bf as\_ld\_yml() }{- turn a computational component list into YAML with class information}
\item{\bf ld\_cc\_dendro() }{- create a dendrogram from a list of computational components}
\item{\bf ld\_chunk\_opts() }{- apply chunk options to a presentation object}
\item{\bf ld\_ioslides\_header() }{- create an ioslides presentation header}
\item{\bf ld\_make\_chunks() }{- write a listdown object to a string}
\item{\bf ld\_rmarkdown\_header() }{- create an R Markdown header}
\item{\bf ld\_workflowr\_header() }{- create a worflowr header}
\item{\bf listdown() }{- create a listdown object to create an R Markdown document}
\end{itemize}

@@

Suppose we have just completed and analysis and have collected all of
the results into a list where the list elements are roughly in the order
we would like to present them in a document. It may be noted that this
is not always how computational components derived from data analyses
are collated. Often individual components are stored in multiple
locations on a single machine or across machines. However, it is
important to realize that even for analyses on large-scale data, the
digital artifacts to be presented are relatively small. Centralizing
them makes it easier to access them, since they don't need to be found
in mulitple locations. Also, storing them as a list provides a
hierarchical structure that translates directly to a document as we will
see below.

As a first example, we will consider the a list of visualizations from
the anscombe data set. The list is composed of four \pkg{ggplot2}
\citep{wickham2016}elements (named Linear, Non Linear, Outlier Vertical,
and Outlier Horizontal) each containing a scatter plot from the famous
Anscome Quartet - made available in the \pkg{datasets} package
\citep{R}. From the \texttt{computational\_components} list, we would
like to create a document with four sections with names corresponding to
the list names, each containing their respective visualizations.

@@

\begin{CodeInput}
R> # Use ggplot2 to create the visualizations.
R> library(ggplot2)
R> 
R> # Load the Anscombe Quartet.
R> data(anscombe)
R> 
R> # Create the ggplot objects to display.
R>   computational_components <- list(
R+     Linear = ggplot(anscombe, aes(x = x1, y = y1)) + geom_point(),
R+     `Non Linear` = ggplot(anscombe, aes(x = x2, y = y2)) + geom_point(),
R+     `Outlier Vertical`= ggplot(anscombe, aes(x = x3, y = y3)) + 
R+       geom_point(),
R+     `Outlier Horizontal` =  ggplot(anscombe, aes(x = x4, y = y4)) + 
R+       geom_point())
R> 
R> # Save the file to disk to be read by the output R Markdown document.
R> saveRDS(computational_components, "comp-comp.rds")
\end{CodeInput}
\end{CodeChunk}

@@

Creating a document from the \texttt{computational\_components} requires
two steps. First, we will create a \texttt{listdown} object that
specifies how the \texttt{computational\_components} object will be
loaded into the document, which libraries and code needs to be included,
and how the list elements will be presented in the output R markdown
document.

@@

\begin{CodeInput}
R> library(listdown)
R> 
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"),
R+                package = "ggplot2")
\end{CodeInput}
\end{CodeChunk}

The \texttt{ld} object, along with the computational components in the
\texttt{comp-comp.rds} file are sufficient to to create the sections,
subsections, and R chunks of a document. The only other thing requires
to create the document is the header. The listdown package currently
supports regular R Markdown and \pkg{workflowr} as \code{yml} objects
from the \pkg{yaml} package \citep{yaml}. These objects are stored as
named lists in R and are easily modified to accomodate document
parameters. A complete document can then be written to the console using
the code shown below. It could easily be written to a file for rendering
using the \texttt{writeLines()} function, for example.

@@

\begin{CodeInput}
R> doc <- c(
R>   as.character(ld_rmarkdown_header("Anscombe's Quartet",
R>                                    author = "Francis Anscombe",
R>                                    date = "1973")),
R>   ld_make_chunks(ld))
R> 
R> # Write the document.
R> writeLines("anscome-example.rmd")
R> 
R> # Show the document.
R> doc
\end{CodeInput}
\end{CodeChunk}

@@

The \texttt{listdown()} function provides \emph{document-wide} R chunk
options for displaying computational components. The chunk options are
exactly the same as those in the R markdown document and can be used to
tailor the default presentation for a variety of needs. The complete set
of options can be found in the R Markdown Reference Guide
\citep{rmarkdownref}. As a concrete example, the code used to create
present the plots could be hidden in the output document using the
following code.

@@

\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
R+                package = "ggplot2",
R+                echo = FALSE)
R> 
R> # Show the updated chunk where the library is loaded and data is read.
R> ld_make_chunks(ld)[1:7]
\end{CodeInput}

@@

The first example is simple in part because the ggplot objects both
contains the data we want to display and, at the same time, provides the
mechanism for presenting them - rendering them in a graph. However, this
is not always the case. The objects being stored in the list of
computational components may not translate directly to the presentation
in a document. In these cases, a function is needed that takes the list
component and returns an object to be displayed. For example, suppose
that, along with showing graphs from the Anscombe Quartet, we would like
to include the data themselves. We could add the data to the
\texttt{computational\_components} list and then create the document
with:

@@

\begin{CodeOutput}
[1] "# Data"              
[2] ""                    
[3] "```{r echo = FALSE}"
[4] "cc_list$Data"        
[5] "```"                
\end{CodeOutput}
\end{CodeChunk}

In this case, the \pkg{listdown} package will show the entire data set
as is the default specified. However, suppose we do not want to show the
entire data set in the document. This is common, especially when the
data set is large and requires too much vertical space in the outputted
document resulting in too much or irrelevant data being shown. Instead,
we would like to output to an html document where the data is shown in a
\texttt{datatable} thereby controlling the amount of real-estate needed
to present the data and, at the same time, providing the user with
interactivity to sort and search the data set.

In \pkg{listdown}, a function or method that implements the presentation
of a computational component is referred to as a \emph{decorator} since
if follows the classic decorator pattern described in \citet{gamma1995}.
A decorator takes the element that will be presented as an argument and
returns an object for presentation in the output directory. A decorator
is specified using the \texttt{decorator} parameter of the
\texttt{listdown()} function using a named list where the name
corresponds to the type and the element correspond to the function or
method that will decorate an object of that type. For example, the
\texttt{anscombe} data set can be decorated with the
\texttt{DT::datatable()} function \citep{xie2020} as:

@@

\begin{CodeOutput}
[1] "# Data"                  
[2] ""                       
[3] "```{r}"                  
[4] "datatable(cc_list$Data)"
[5] "```"                    
\end{CodeOutput}
\end{CodeChunk}

List names in the \texttt{decorator} argument provide a key to which a
function or method is mapped. The underlying decorator resolution is
implemented for a given computational component by going through
decorator names sequentially to see if the component inherits from the
name using the \texttt{inherits()} function. The function or method is
selected from the corresponding name which the element first inherits
from. This means that when customizing the presentation of objects that
inherit from a common class, the more abstract classes should appear at
the end of the list. This will ensure that specialized classes will be
encountered first in the resolution process. It should be noted that an
object's type is first checked against the decorator name list and then
checked to see if it is a list. This allows a user to both decorate a
list and retain \texttt{"list"} in its class attributes.

A separate argument, \texttt{default\_decorator}, allows the user to
specify the default decorator for an object whose type does not appear
in the \texttt{decorator} list. This allows the user to specify any
class name for the decorator and avoids a potential type name collision
with a default decorator whose name is determined by convention. By
default, this argument is set to \texttt{identity} but it can be use to
not display a computational component by default if the argument is set
to \texttt{NULL}.

@@

A \pkg{listdown} object specifies the location of a list of
computational components and options for presenting those components in
an R Markdown document. The list is a hierarchical data structure that
also provides the structure of the outputted document. A corresponding
document has two sections ``Iris'' and ``Sepal.Length''. The latter has
three subsections ``Sepal.Width'', ``Petal.Length'', and ``Colored''.
The ``Colored'' subsection has two sub-subsections, ``Sepal.Width'' and
``Petal.Length''. The structure can be seen using the
\texttt{ld\_cc\_dendro()} function.

@@

\begin{CodeInput}
R> # Create a more hierarchical list of computational components.
R> comp_comp2 <- list(
R+   Iris = iris,
R+   Sepal.Length = list(
R+     Sepal.Width = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
R+       geom_point(),
R+     Petal.Length = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
R+       geom_point(),
R+     Colored = list(
R+       Sepal.Width = ggplot(iris, 
R+                           aes(x = Sepal.Length, y = Sepal.Width, 
R+                               color = Species)) + geom_point(),
R+       Petal.Length = ggplot(iris,
R+                             aes(x = Sepal.Length, y = Petal.Length, 
R+                                 color = Species)) + geom_point())))
R> 
R> # Create the dendrogram.
R> ld_cc_dendro(comp_comp2)
\end{CodeInput}

@@

Both the \texttt{ld\_cc\_dendro()} and \texttt{ld\_make\_chunks()}
functions work by recursively descending the computational components
list depth-first. If the list containing and element has a name, it is
written to the output as a section, subsection, subsubsection, etc. to a
return string. If the visited list element is itself a list, then the
same procedure is called on the child list through a recursive call. If
the element is not a list, then it is outputted inside an R Markdown
chunk in the return string using the appropropriate decorator.

@@

The \texttt{listdown()} function facilitates the insertion of
initialization code through the \texttt{init\_expr} argument. Code is
inserted immediately after the libraries are loaded in the R Markdown
document. In general, it is suggested that the number of initial
expressions be kept small so that the \proglang{R} Markdown document is
easy to read. If a large number of functions are required by the target
\proglang{R} Markdown document then they can be put into a file and
sourced using the initial expression. As an example, suppose we are
creating an html document and presenting data using the
\texttt{datatable()} function. However, we do not want to include the
search capabilities provided. This can be easily accomplished by
creating a new function, \texttt{datatable\_no\_search()}, created using
the \texttt{partial()} function \cite{purrr} to partially apply
\texttt{list(dom\ =\ \textquotesingle{}t\textquotesingle{})} to the
\texttt{options} argument of \texttt{datatable}.

@@

The document-wide chunk option specification provides the default chunk
options for output documents generated using \pkg{listdown}. However,
the presentation of a data object often varies by type. For example, we
may want to specify the height and width of a graph, but not a table.
This is accomplished in the \pkg{listdown} package when a
\texttt{listdown} object is created using the
\texttt{decorator\_chunk\_opts} option in the \texttt{listdown()}
function. For example, associating all \texttt{ggplot} objects with
\proglang{R} chunks having a width of 100 and a height of 200 can be
accomplished with the following code and it can be seen that only chunk
options associated with a plot are modified.

@@

Along with providing decorator-wide chunk options, it is also possible
to control individual chunk options. The capability is distinct from the
document-wide and decorator-wide specification of options in that it
must be applied to the computational component list element whose
associated options will be modified, rather than the \texttt{listdown}
object. This is because the \texttt{listdown} only specifies how classes
of objects should be presented. To modify the chunk options associated
with a specific list element the list element is provided with a set of
attributes that can be queried by the \texttt{ld\_chunk\_opts()}
function as the output document is being generated. Because of the ad
hoc nature of this capability, its use is discouraged. A better
solution, that maintains the behavior is to add class information to the
list element and specify decorator-wide chunk options for the new class.
This maintains the separation of the computational component list, which
maintains the document structure and data for presentation from the
specification of how the document will be created and rendered.

@@

\begin{CodeOutput}
[1] "# Iris"             
[2] ""                   
[3] "```{r echo = TRUE}"
[4] "cc_list$Iris"       
[5] "```"               
\end{CodeOutput}
\end{CodeChunk}

@@

As mentioned before, we have found the \pkg{listdown} package
particularly helpful for reporting the results of clinical trials
thereby creating a basis for discussion and collaboration between
(bio)statisticians and the clinicians running the trials. For this use
case, the context and data collection procedures are well-understood and
as a result very few narrative components are needed. It is also the
case that the modes of presentation (tables, scatterplots,
survivalplots, consort diagrams, etc.) are standardized. The goal in
presenting the trial characteristics is to identify problems in the
data, monitor trial enrollment and response, quantify known
relationships among the data, and test hypotheses about a therapy's
efficacy.

In practice we generally separate the data cleaning, exploration,
analysis, monitoring, and presentation components. The computational
component list has tens of elements with thousands of visualizations.
These facts, coupled with the privacy constraints make a complete
example difficult for the purposes of this paper. So, below we provide a
simple example of the types of documents we provide using the
\texttt{gtsummary::trial} dataset. While it is not complete, it does
convey they types of reports we are currently generating for completed
and ongoing clinical trials.

@@

\begin{CodeInput}
R> library(gtsummary)
R> library(dplyr)
R> library(survival)
R> library(survminer)
R> library(rmarkdown)
R> 
R> # Function to create the computational components.
R> make_surv_cc <- function(trial, treat, surv_cond_chars) {
R+   # Create Table 1 by treatment
R+   table_1 <- trial %>%
R+     tbl_summary(by = all_of(treat)) %>%
R+     gtsummary::as_flextable()
R+ 
R+   # Create survival plots base on survival characteristics
R+   scs <- lapply(c("1", surv_cond_chars),
R+                 function(sc) {
R+                   sprintf("Surv(ttdeath, death) ~ %s + %s", treat, sc) %>%
R+                     as.formula() %>%
R+                     surv_fit(trial) %>%
R+                     ggsurvplot()
R+                 })
R+   names(scs) <- c("Overall", tools::toTitleCase(surv_cond_chars))
R+   list(`Table 1` = table_1, `Survival Plots` = scs)
R+ }
R> 
R> # Create the computational components and write to surv-cc.rds.
R> surv_cc <- make_surv_cc(trial, treat = "trt",
R+                         surv_cond_chars = c("stage", "grade"))
R> 
R> ld_cc_dendro(surv_cc)
\end{CodeInput}

@@

As shown before, the report is created by saving the computational
components, creating a listdown object, writing the \proglang{R}
Markdown document, and rendering it. The resulting document,
trial-report.html, can then be placed in a shared space where it can be
viewed and interpreted by stakeholders in the clincial trial. The
\proglang{R} Markdown document created by this code is shown in Appendix
1.

@@

\bibliography{references}

\section*{Appendix 1: R Markdown produced in the simple example}
\label{appendix}

>>>>

\author{
Michael Kane\\Yale University \And Xun (Tony) Jiang\\Amgen
Inc. \And Simon Urbanek\\The University of Aukland
}
\title{On the Programmatic Generation of Reproducible Documents}

@@

      Xun (Tony) Jiang\\
  Amgen Inc.\\
  One Amgen Center DriveThousand Oaks, CA 91320-1799, USA\\
  E-mail: \href{mailto:xunj@amgen.com}{\nolinkurl{xunj@amgen.com}}\\

@@

One area where we have found \pkg{listdown} is particularly useful is in
the reporting and research of clinical trial data. These collaborations
tend to be between (bio)statisticians and clinicians either analyzing
past trial data to formulate a new trial or in trial monitoring where
trial telemetry (enrollment, responses, etc.) is reported and initial
analyses are conveyed to a clinician. The associated presentations
require very little context - clinicians often have as good an
understanding of the data collected as that of the statistician's -
often eliminating or significantly reducing the need for narrative
components. At the same time, a large number of hierarchical,
heterogeneous artifacts (tables and multiple types of plots) need to be
conveyed thereby making the manual creation of \proglang{R} Markdown
documents inconvenient.

In this case, data presentation can be fixed across trials. This is
especially true in the initial stages, which focus on patient
demographics and enrollment. This approach has made it convenient for
our group to quickly generate standardize and complete reports for
multiple trials concurrently. To date, we have used listdown to report
on five clinical trials, with another two currently in process. Results
are disseminated using the \pkg{workflowr} package, usually with nine
tabs conveying aspects of the data from collection through several
different analyses, and each tab containing approximately five to thirty
tables, plots, or other artifacts including \pkg{trelliscopejs}
\citep{trelliscopejs} environments which may hold hundreds of graphs. By
generating many presentation artifacts we are able to address data-drive
questions and issues during collaborative sessions and by carefully
structuring these elements we allowing all members to participate in the
process.

@@

\begin{itemize}
\item{\bf as\_ld\_yml() }{- turn a computational component list into YAML with class information}
\item{\bf ld\_cc\_dendro() }{- create a dendrogram from a list of computational components}
\item{\bf ld\_chunk\_opts() }{- apply chunk options to a presentation object}
\item{\bf ld\_ioslides\_header() }{- create an ioslides presentation header}
\item{\bf ld\_make\_chunks() }{- write a listdown object to a string}
\item{\bf ld\_rmarkdown\_header() }{- create an R Markdown header}
\item{\bf ld\_workflowr\_header() }{- create a worflowr header}
\item{\bf listdown() }{- create a listdown object to create an \proglang{R} Markdown document}
\item{\bf print.listdown()}{- print the listdown options for \proglang{R} Markdown document creation}
\end{itemize}

@@

Suppose we have just completed and analysis and have collected all of
the results into a list where the list elements are roughly in the order
we would like to present them in a document. It may be noted that this
is not always how computational components derived from data analyses
are collated. Often individual components are stored in multiple
locations on a single machine or across machines. However, it is
important to realize that even for analyses on large-scale data, the
digital artifacts to be presented are relatively small. Centralizing
them makes it easier to access them, since they don't need to be found
in multiple locations. Also, storing them as a list provides a
hierarchical structure that translates directly to a document as we will
see below.

As a starting point, we will consider the a list of visualizations from
the Anscombe data set below. The list is composed of four \pkg{ggplot2}
\citep{wickham2016} elements (named Linear, Non Linear, Outlier
Vertical, and Outlier Horizontal) each containing a scatter plot from
the Anscombe Quartet - made available in the \pkg{datasets} package
\citep{R} as shown in Figure \ref{cc1}. From the
\code{computational_components} list, we would like to create a document
with four sections with names corresponding to the list names, each
containing their respective visualizations. The structure of a document
derived from the \code{computational_components} list can be visualized
using the \code{ld_cc_dendro()} function, and its output is below.

@@

\begin{CodeInput}
R> library(ggplot2)
R> 
R> library(listdown)
R> 
R> data(anscombe)
R> 
R> computational_components <- list(
R+   Linear = ggplot(anscombe, aes(x = x1, y = y1)) + geom_point(),
R+   `Non Linear` = ggplot(anscombe, aes(x = x2, y = y2)) + geom_point(),
R+   `Outlier Vertical`= ggplot(anscombe, aes(x = x3, y = y3)) + 
R+     geom_point(),
R+   `Outlier Horizontal` =  ggplot(anscombe, aes(x = x4, y = y4)) + 
R+     geom_point())
R> 
R> ld_cc_dendro(computational_components)
\end{CodeInput}

computational_components
  |-- Linear
  |  o-- object of type(s):gg ggplot
  |-- Non Linear
  |  o-- object of type(s):gg ggplot
  |-- Outlier Vertical
  |  o-- object of type(s):gg ggplot
  o-- Outlier Horizontal
     o-- object of type(s):gg ggplot
\end{CodeOutput}
\end{CodeChunk}

@@

Creating a document whose structure and content are described
\code{computational_components} requires two steps. First, we will
create a \code{listdown} object specifying how the \newline
\code{computational_components} object will be loaded into the document,
which libraries and code needs to be included, and how the list elements
will be presented in the output R markdown document. A human-readable
\code{print} function is included in the package and is the default
output of the object. It should be noted that the output shows options
that will be described and illustrated later.

@@

\begin{CodeInput}
R> saveRDS(computational_components, "comp-comp.rds")
R> 
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"),
R+                package = "ggplot2")
R> 
R> ld
\end{CodeInput}

Listdown object description

Package(s) imported:
    ggplot2

Initial expression(s) (run after packages are loaded):
    (none)

Expression to read data:
    readRDS("comp-comp.rds")

Decorator(s):
    (none)

Defaut decorator:
    identity

Chunk option(s):
    (none)

Decorator chunk option(s):
    (none)
\end{CodeOutput}
\end{CodeChunk}

The \code{ld} object, along with the computational components in the
\code{comp-comp.rds} file are sufficient to to create the sections,
subsections, and \proglang{R} chunks of a document. The only other thing
requires to create the document is the header. The listdown package
currently supports regular R Markdown and \pkg{workflowr} as \code{yml}
objects from the \pkg{yaml} package \citep{yaml}. These objects are
stored as named lists in \proglang{R} and are easily modified to
accommodate document parameters. A complete document can then be written
to the console using the code shown below. It could easily be written to
a file for rendering using the \code{writeLines()} function, for
example.

@@

\begin{CodeInput}
R> doc <- c(
R>   as.character(ld_rmarkdown_header("Anscombe's Quartet",
R>                                    author = "Francis Anscombe",
R>                                    date = "1973")),
R>   ld_make_chunks(ld))
R> 
R> writeLines("anscome-example.rmd")
R> 
R> doc
\end{CodeInput}
\end{CodeChunk}

@@

The \code{listdown()} function provides \emph{document-wide} R chunk
options for displaying computational components. The chunk options are
exactly the same as those in the R markdown document and can be used to
tailor the default presentation for a variety of needs. The complete set
of options can be found in the R Markdown Reference Guide
\citep{rmarkdownref}. As a concrete example, the code used to create
present the plots could be hidden in the output document using the
following code.

@@

\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
R+                package = "ggplot2",
R+                echo = FALSE)
R> 
R> ld_make_chunks(ld)[1:7]
\end{CodeInput}

@@

The first example is simple in part because the ggplot objects both
contains the data we want to display and, at the same time, provides the
mechanism for presenting them - rendering them in a graph. However, this
is not always the case. The objects being stored in the list of
computational components may not translate directly to the presentation
in a document. In these cases, a function is needed that takes the list
component and returns an object to be displayed. For example, suppose
that, along with showing graphs from the Anscombe Quartet, we would like
to include the data themselves. We could add the data to the
\code{computational_components} list and then create the document with:

@@

\begin{CodeOutput}
[1] "# Data"              ""                    "```{r echo = FALSE}"
[4] "cc_list$Data"        "```"                
\end{CodeOutput}
\end{CodeChunk}

In this case, the \pkg{listdown} package will show the entire data set
as is the default specified. However, suppose we do not want to show the
entire data set in the document. This is common, especially when the
data set is large and requires too much vertical space in the outputted
document resulting in too much or irrelevant data being shown. Instead,
we would like to output to an html document where the data is shown in a
\code{datatable} thereby controlling the amount of real-estate needed to
present the data and, at the same time, providing the user with
interactivity to sort and search the data set.

In \pkg{listdown}, a function or method that implements the presentation
of a computational component is referred to as a \emph{decorator} since
if follows the classic decorator pattern described in \citet{gamma1995}.
A decorator takes the element that will be presented as an argument and
returns an object for presentation in the output directory. A decorator
is specified using the \code{decorator} parameter of the
\code{listdown()} function using a named list where the name corresponds
to the type and the element correspond to the function or method that
will decorate an object of that type. For example, the \code{anscombe}
data set can be decorated with the \code{DT::datatable()} function
\citep{xie2020} as:

@@

\begin{CodeOutput}
[1] "# Data"                  ""                       
[3] "```{r}"                  "datatable(cc_list$Data)"
[5] "```"                    
\end{CodeOutput}
\end{CodeChunk}

List names in the \code{decorator} argument provide a key to which a
function or method is mapped. The underlying decorator resolution is
implemented for a given computational component by going through
decorator names sequentially to see if the component inherits from the
name using the \code{inherits()} function. The function or method is
selected from the corresponding name which the element first inherits
from. This means that when customizing the presentation of objects that
inherit from a common class, the more abstract classes should appear at
the end of the list. This will ensure that specialized classes will be
encountered first in the resolution process. It should be noted that an
object's type is first checked against the decorator name list and then
checked to see if it is a list. This allows a user to both decorate a
list and retain \code{"list"} in its class attributes.

A separate argument, \code{default_decorator}, allows the user to
specify the default decorator for an object whose type does not appear
in the \code{decorator} list. This allows the user to specify any class
name for the decorator and avoids a potential type name collision with a
default decorator whose name is determined by convention. By default,
this argument is set to \code{identity} but it can be use to not display
a computational component by default if the argument is set to
\code{NULL}.

@@

A \pkg{listdown} object specifies the location of a list of
computational components and options for presenting those components in
an R Markdown document. The list is a hierarchical data structure that
also provides the structure of the outputted document. A corresponding
document has two sections ``Iris'' and ``Sepal.Length''. The latter has
three subsections ``Sepal.Width'', ``Petal.Length'', and ``Colored''.
The ``Colored'' subsection has two subsubsections, ``Sepal.Width'' and
``Petal.Length''. The structure can once again be seen using the
\code{ld_cc_dendro()} function.

@@

\begin{CodeInput}
R> comp_comp2 <- list(
R+   Iris = iris,
R+   Sepal.Length = list(
R+     Sepal.Width = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
R+       geom_point(),
R+     Petal.Length = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
R+       geom_point(),
R+     Colored = list(
R+       Sepal.Width = ggplot(iris, 
R+                           aes(x = Sepal.Length, y = Sepal.Width, 
R+                               color = Species)) + geom_point(),
R+       Petal.Length = ggplot(iris,
R+                             aes(x = Sepal.Length, y = Petal.Length, 
R+                                 color = Species)) + geom_point())))
R> 
R> ld_cc_dendro(comp_comp2)
\end{CodeInput}

@@

Both the \code{ld_cc_dendro()} and \code{ld_make_chunks()} functions
work by recursively descending the computational components list
depth-first. If the list containing and element has a name, it is
written to the output as a section, subsection, subsubsection, etc. to a
return string. If the visited list element is itself a list, then the
same procedure is called on the child list through a recursive call. If
the element is not a list, then it is outputted inside an R Markdown
chunk in the return string using the appropriate decorator.

@@

The \code{listdown()} function facilitates the insertion of
initialization code through the \code{init_expr} argument. Code is
inserted immediately after the libraries are loaded in the R Markdown
document. In general, it is suggested that the number of initial
expressions be kept small so that the \proglang{R} Markdown document is
easy to read. If a large number of functions are required by the target
\proglang{R} Markdown document then they can be put into a file and
sourced using the initial expression. As an example, suppose we are
creating an html document and presenting data using the
\code{datatable()} function. However, we do not want to include the
search capabilities provided. This can be easily accomplished by
creating a new function, \code{datatable_no_search()}, created using the
\code{partial()} function \cite{purrr} to partially apply
\code{list(dom = 't')} to the \code{options} argument of
\code{datatable}.

@@

The document-wide chunk option specification provides the default chunk
options for output documents generated using \pkg{listdown}. However,
the presentation of a data object often varies by type. For example, we
may want to specify the height and width of a graph, but not a table.
This is accomplished in the \pkg{listdown} package when a
\code{listdown} object is created using the \code{decorator_chunk_opts}
option in the {]}code\{listdown()\} function. For example, associating
all \code{ggplot} objects with \proglang{R} chunks having a width of 100
and a height of 200 can be accomplished with the following code and it
can be seen that only chunk options associated with a plot are modified.

@@

Along with providing decorator-wide chunk options, it is also possible
to control individual chunk options. The capability is distinct from the
document-wide and decorator-wide specification of options in that it
must be applied to the computational component list element whose
associated options will be modified, rather than the \code{listdown}
object. This is because the \code{listdown} only specifies how classes
of objects should be presented. To modify the chunk options associated
with a specific list element the list element is provided with a set of
attributes that can be queried by the \code{ld_chunk_opts()} function as
the output document is being generated. Because of the ad hoc nature of
this capability, its use is discouraged. A better solution, that
maintains the behavior is to add class information to the list element
and specify decorator-wide chunk options for the new class. This
maintains the separation of the computational component list, which
maintains the document structure and data for presentation from the
specification of how the document will be created and rendered.

@@

\begin{CodeOutput}
[1] "# Iris"             ""                   "```{r echo = TRUE}"
[4] "cc_list$Iris"       "```"               
\end{CodeOutput}
\end{CodeChunk}

@@

As mentioned before, we have found the \pkg{listdown} package
particularly helpful for reporting the results of clinical trials
thereby creating a basis for discussion and collaboration between
(bio)statisticians and the clinicians running the trials. For this use
case, the context and data collection procedures are well-understood and
as a result very few narrative components are needed. It is also the
case that the modes of presentation (tables, scatterplots, survival
plots, consort diagrams, etc.) are standardized. The goal in presenting
the trial characteristics is to identify problems in the data, monitor
trial enrollment and response, quantify known relationships among the
data, and test hypotheses about a therapy's efficacy.

In practice we generally separate the data cleaning, exploration,
analysis, monitoring, and presentation components. The computational
component list has tens of elements with thousands of visualizations.
These facts, coupled with the privacy constraints make a complete
example difficult for the purposes of this paper. So, below we provide a
simple example of the types of documents we provide using the
\code{gtsummary::trial} data set. While it is not complete, it does
convey they types of reports we are currently generating for completed
and ongoing clinical trials.

@@

\begin{CodeInput}
R> library(gtsummary)
R> library(dplyr)
R> library(survival)
R> library(survminer)
R> library(rmarkdown)
R> 
R> make_surv_cc <- function(trial, treat, surv_cond_chars) {
R+   table_1 <- trial %>%
R+     tbl_summary(by = all_of(treat)) %>%
R+     gtsummary::as_flextable()
R+ 
R+   scs <- lapply(c("1", surv_cond_chars),
R+                 function(sc) {
R+                   sprintf("Surv(ttdeath, death) ~ %s + %s", treat, sc) %>%
R+                     as.formula() %>%
R+                     surv_fit(trial) %>%
R+                     ggsurvplot()
R+                 })
R+   names(scs) <- c("Overall", tools::toTitleCase(surv_cond_chars))
R+   list(`Table 1` = table_1, `Survival Plots` = scs)
R+ }
R> 
R> surv_cc <- make_surv_cc(trial, treat = "trt",
R+                         surv_cond_chars = c("stage", "grade"))
R> 
R> ld_cc_dendro(surv_cc)
\end{CodeInput}

@@

As shown before, the report is created by saving the computational
components, creating a \code{listdown} object, writing the \proglang{R}
Markdown document, and rendering it. The resulting document,
trial-report.html, can then be placed in a shared space where it can be
viewed and interpreted by stakeholders in the clinical trial. The
\proglang{R} Markdown document created by this code is shown in
supplementary materials.

@@

\bibliography{../references}
